Towards a free/open-source universal-dependency treebank for Kazakh

### 2015-08-27 2015-turklang-kazdep/paper.tex
another fix to table 4
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@61594 72bbbca6-d526-0410-a7d9-f06f51895060
cop → _, not .
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@61596 72bbbca6-d526-0410-a7d9-f06f51895060
update numbers!
git-svn-id: https://svn.code.sf.net/p/apertium/svn/branches/papers/@61609 72bbbca6-d526-0410-a7d9-f06f51895060

\begin{table}[htbp]
	\centering
	\caption{The functions associated with each case in Kazakh as currently found in the corpus.}
	\label{tab:casefunc}
	\begin{tabular}{lrrrrrrrr}
		\toprule
			 & \udtag{iobj} & \udtag{cmpnd} & \udtag{nmod} & \udtag{nmod:poss} & \udtag{dobj} & \udtag{root} & \udtag{subj} & \udtag{voc} \\
		\midrule
			\gmk{nom} & — & 1 & 199 & 157 & 72 & 26 & 279 & 1 \\
			\gmk{acc} & — & — & 1 & — & 101 & — & — & — \\
			\gmk{gen} & — & — & 120 & 120 & — & — & 12 & — \\
			\gmk{dat} & 1 & — & 113 & — & 13 & — & — & — \\
			\gmk{loc} & — & — & 113 & — & — & 6 & — & — \\
			\gmk{abl} & — & — & 64 & — & — & — & — & — \\
			\gmk{ins} & — & — & 33 & — & — & — & — & — \\
		\bottomrule
	\end{tabular}
\end{table}
\vspace{-0.75em}

@@

        \begin{subfigure}[b]{0.25\textwidth}
                \centering
                \begin{dependency}[theme = simple, font = \small]
                   \begin{deptext}[column sep=0.2cm]
                                Айгүл \& студент \& · \\
                                \gmk{np} \& \gmk{n} \& \gmk{aor} \\
                                \tgloss{Aygül} \& \tgloss{student} \& \tgloss{is} \\
                        \end{deptext}
                        \deproot[edge unit distance=1.3ex]{2}{\udlabel{root}}
                        \depedge{2}{1}{\udlabel{subj}}
                        \depedge{2}{3}{\udlabel{cop}}
                \end{dependency}
                \caption{Non-surfaced copula.}\label{fig:cop1}
        \end{subfigure}
        \quad
        \begin{subfigure}[b]{0.25\textwidth}
                \centering
                \begin{dependency}[theme = simple, font = \small]
                   \begin{deptext}[column sep=0.2cm]
                                Айгүл \& студент \& болған \\
                                \gmk{np} \& \gmk{n} \& \gmk{past} \\
                                \tgloss{Aygül} \& \tgloss{student} \& \tgloss{was}\\
                        \end{deptext}
                        \deproot[edge unit distance=1.3ex]{2}{\udlabel{root}}
                        \depedge{2}{1}{\udlabel{subj}}
                        \depedge{2}{3}{\udlabel{cop}}
                \end{dependency}
                \caption{Copula in past tense.}\label{fig:cop2}
        \end{subfigure}
        \quad
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \begin{dependency}[theme = simple, font = \small]
                   \begin{deptext}[column sep=0.2cm]
                                Айгүл \& оның \& қайда \& екенін \& білмейді. \\
                                \gmk{np} \& \gmk{gen} \& \gmk{adv} \& \gmk{cop} \& \gmk{neg.aor} \\
                                \tgloss{Aygül} \& \tgloss{his} \& \tgloss{where} \& \tgloss{being} \& \tgloss{knows not} \\
                        \end{deptext}
                        \deproot[edge unit distance=1.3ex]{5}{\udlabel{root}}
                        \depedge{3}{2}{\udlabel{subj}}
                        \depedge{5}{3}{\udlabel{ccomp}}
                        \depedge{3}{4}{\udlabel{cop}}
                \end{dependency}
                \caption{Copula in an embedded clause.}\label{fig:cop3}
        \end{subfigure}
        \caption{Dependency trees of copula constructions.}
\end{figure}
\vspace{-0.75em}

@@

In order to test the utility of the treebank in a real-world setting, we trained
and evaluated a number of models using the popular MaltParser tool \parencite{nivre07}.
MaltParser is a toolkit for data-driven dependency parsing; it can learn a parsing
model from treebank data and apply this model to parse unseen sentences. The parser
has a large number of options and parameters that need to be optimised. 
To select the best parser configuration we relied on 
MaltOptimiser \parencite{ballesteros15}. The optimiser was run separately for each of 
the model configurations.

@@

\begin{table}[htbp]
	\caption{Preliminary parsing results from MaltParser using different models. The numbers in brackets denote the upper and lower bounds found during cross-validation.}
	\centering
	\begin{tabular}{llrr}
		\toprule
			\textbf{Features}       & \textbf{Algorithm} &\textbf{LAS} [range] & \textbf{UAS} [range] \\
		\midrule
			surface                & nivreeager  & 33.4 [23.4, 42.5] & 50.9 [41.3, 63.2] \\
			surface+lemma          & nivreeager  & 33.4 [23.4, 42.5] & 50.9 [41.3, 63.2] \\
			surface+lemma+POS      & nivreeager  & 55.2 [43.4, 75.0] & 73.7 [65.8, 92.1] \\
			surface+lemma+POS+MSD  & nivreeager  & 55.2 [43.4, 75.0] & 73.7 [65.8, 92.1] \\
		\bottomrule
	\end{tabular}
	\label{table:eval}
\end{table}

@@

The results we obtain, shown in table~\ref{table:eval}, are similar to those obtained with similar 
sized treebanks, for example the Irish treebank of \textcite{Lynn12}, for which an LAS of
63.3 and a UAS of 73.3 were reported with the best model.  Adding structural features to the model substantially improves the performance of the parser, although we find no improvement in using full morphosyntactic description (MSD) over using simply the first part-of-speech (POS) tag.

@@

>>>>

\begin{table}[htbp]
	\centering
	\caption{The functions associated with each case in Kazakh as currently found in the corpus.}
	\label{tab:casefunc}
	\begin{tabular}{lrrrrrrrr}
		\toprule
			 & \udtag{iobj} & \udtag{cmpnd} & \udtag{nmod} & \udtag{nmod:poss} & \udtag{dobj} & \udtag{root} & \udtag{subj} & \udtag{voc} \\
		\midrule
			\gmk{nom} & — & 1 & 199 & 157 & 72 & 26 & 279 & 1 \\
			\gmk{acc} & — & — & — & — & 102 & — & — & — \\
			\gmk{gen} & — & — & 120 & 120 & — & — & 12 & — \\
			\gmk{dat} & 1 & — & 113 & — & 13 & — & — & — \\
			\gmk{loc} & — & — & 113 & — & — & 6 & — & — \\
			\gmk{abl} & — & — & 64 & — & — & — & — & — \\
			\gmk{ins} & — & — & 33 & — & — & — & — & — \\
		\bottomrule
	\end{tabular}
\end{table}
\vspace{-0.75em}

@@

        \begin{subfigure}[b]{0.25\textwidth}
                \centering
                \begin{dependency}[theme = simple, font = \small]
                   \begin{deptext}[column sep=0.2cm]
                                Айгүл \& студент \& \_ \\
                                \gmk{np} \& \gmk{n} \& \gmk{aor} \\
                                \tgloss{Aygül} \& \tgloss{student} \& \tgloss{is} \\
                        \end{deptext}
                        \deproot[edge unit distance=1.3ex]{2}{\udlabel{root}}
                        \depedge{2}{1}{\udlabel{subj}}
                        \depedge{2}{3}{\udlabel{cop}}
                \end{dependency}
                \caption{Non-surfaced copula.}\label{fig:cop1}
        \end{subfigure}
        \quad
        \begin{subfigure}[b]{0.25\textwidth}
                \centering
                \begin{dependency}[theme = simple, font = \small]
                   \begin{deptext}[column sep=0.2cm]
                                Айгүл \& студент \& болған \\
                                \gmk{np} \& \gmk{n} \& \gmk{past} \\
                                \tgloss{Aygül} \& \tgloss{student} \& \tgloss{was}\\
                        \end{deptext}
                        \deproot[edge unit distance=1.3ex]{2}{\udlabel{root}}
                        \depedge{2}{1}{\udlabel{subj}}
                        \depedge{2}{3}{\udlabel{cop}}
                \end{dependency}
                \caption{Copula in past tense.}\label{fig:cop2}
        \end{subfigure}
        \quad
        \begin{subfigure}[b]{0.40\textwidth}
                \centering
                \begin{dependency}[theme = simple, font = \small]
                   \begin{deptext}[column sep=0.2cm]
                                Айгүл \& оның \& қайда \& екенін \& білмейді. \\
                                \gmk{np} \& \gmk{gen} \& \gmk{adv} \& \gmk{cop} \& \gmk{neg.aor} \\
                                \tgloss{Aygül} \& \tgloss{his} \& \tgloss{where} \& \tgloss{being} \& \tgloss{knows not} \\
                        \end{deptext}
                        \deproot[edge unit distance=1.3ex]{5}{\udlabel{root}}
                        \depedge{3}{2}{\udlabel{subj}}
                        \depedge{5}{3}{\udlabel{ccomp}}
                        \depedge{3}{4}{\udlabel{cop}}
                \end{dependency}
                \caption{Copula in an embedded clause.}\label{fig:cop3}
        \end{subfigure}
        \caption{Dependency trees of copula constructions.}
\end{figure}
\vspace{-0.75em}

@@

In order to test the utility of the treebank in a real-world setting, we trained
and evaluated a number of models using the popular MaltParser tool \parencite{nivre07}.
MaltParser is a toolkit for data-driven dependency parsing; it can learn a parsing
model from treebank data and apply this model to parse unseen sentences. The parser
has a large number of options and parameters that need to be optimised. 
To select the best parser configuration we relied on 
MaltOptimiser \parencite{ballesteros15}. The optimiser was run separately for each of 
the model configurations, and the best combination of parameters was selected.

@@

% system 1: 
% LAS: 0.4285 [0.389+0.406+0.414+0.42+0.427+0.437+0.438+0.445+0.45+0.459]
% UAS: 0.5842 [0.524+0.555+0.572+0.575+0.584+0.584+0.585+0.606+0.628+0.629]

% system 2: 
% nivreeager
% LAS: 0.4285 [0.389+0.406+0.414+0.42+0.427+0.437+0.438+0.445+0.45+0.459]
% UAS: 0.5842 [0.524+0.555+0.572+0.575+0.584+0.584+0.585+0.606+0.628+0.629]

% system 3: 
% nivreeager
% LAS: 0.6499 [0.664+0.649+0.653+0.682+0.617+0.671+0.664+0.625+0.603+0.671]
% UAS: 0.7703 [0.768+0.767+0.763+0.789+0.725+0.778+0.804+0.747+0.762+0.8]

% system 4: 
% nivreeager
% LAS: 0.7688 [0.709+0.725+0.764+0.775+0.776+0.778+0.785+0.786+0.788+0.802]
% UAS: 0.8143 [0.757+0.784+0.809+0.81+0.82+0.823+0.825+0.831+0.832+0.852]

\begin{table}[htbp]
	\caption{Preliminary parsing results from MaltParser using different models. The numbers in brackets denote the upper and lower bounds found during cross-validation.}
	\centering
	\begin{tabular}{llrr}
		\toprule
			\textbf{Features}       & \textbf{Algorithm} &\textbf{LAS} [range] & \textbf{UAS} [range] \\
		\midrule
			surface                & nivreeager  & 42.8 [38.9, 45.9]  & 58.4 [52.4, 62.9] \\
			surface+lemma          & nivreeager  & 42.8 [38.9, 45.9]  & 58.4 [52.4, 62.9]   \\
			surface+lemma+POS      & nivreeager  & 64.9 [66.4, 67.1]  & 77.0 [76.8, 80.0] \\
			surface+lemma+POS+MSD  & nivreeager  & 76.8 [70.9, 80.0]  & 81.4 [75.7, 85.2] \\
		\bottomrule
	\end{tabular}
	\label{table:eval}
\end{table}

@@

The results we obtain, shown in table~\ref{table:eval}, are similar to those obtained with similar 
sized treebanks, for example the Irish treebank of \textcite{Lynn12}, for which an LAS of
63.3 and a UAS of 73.3 were reported with the best model.  Adding structural features to the model 
substantially improves the performance of the parser.

@@

%%%%%%%%%%%%%%%%%%%
%%
%% OLD RESULTS:
